---
title: "Testing_ArchR"
author: "Ravza"
date: "02/07/2024"
output: html_document
---

# ArchR Analysis
Basic tutorial: https://www.archrproject.com/articles/Articles/tutorial.html
Full tutorial: https://www.archrproject.com/bookdown/index.html

## Loading the package
```{r suppressPackageStartupMessages=TRUE, message=FALSE}
library(ArchR)
```

## Getting ready
```{r}
set.seed(1)
addArchRThreads(threads = 16)
addArchRGenome("hg19")
```

## Downloading input materials
```{r}
#inputFile <- "data/atac_v1_pbmc_5k_fragments.tsv.gz"

inputFiles = getTutorialData(tutorial = "hematopoiesis")
#input_file = "/home/imm/grte2681/scripts/HemeFragments/scATAC_BMMC_R1.fragments.tsv.gz"
```

```{r}
inputFiles = getTutorialData(tutorial = "hematopoiesis")
```


## Creating Arrow files
```{r}
# ~ 10 mins
ArrowFiles <- createArrowFiles(
  inputFiles = inputFiles,
  sampleNames = names(inputFiles),
  minTSS = 4, #Don't set this too high because you can always increase later
  minFrags = 1000, 
  addTileMat = TRUE,
  addGeneScoreMat = TRUE
)
```

## If you need to re-start the analysis and have arrow files generated, you can direclty load the arrow files.
```{r}
ArrowFiles = c("scATAC_BMMC_R1.arrow", "scATAC_CD34_BMMC_R1.arrow", "scATAC_PBMC_R1.arrow")
names(ArrowFiles) = c("BMMC", "CD34_BMMC", "PBMC")
```


## Calculating doublets
<simpleError in g$grobs[[legend]]: no such index at level 2>
Incompatible ggplot2 version — Some versions introduce changes in the structure of ggplot objects.
```{r}
# ~ 5 mins
doubScores <- addDoubletScores(
    input = ArrowFiles,
    k = 10, #Refers to how many cells near a "pseudo-doublet" to count.
    knnMethod = "UMAP", #Refers to the embedding to use for nearest neighbor search with doublet projection.
    LSIMethod = 1
)
```

```{r}
doub_bmmc = readRDS("QualityControl/scATAC_BMMC_R1/scATAC_BMMC_R1-Doublet-Summary.rds")
names(doub_bmmc)

ggplot(doub_bmmc$originalDataUMAP, aes(x = X1, y = X2, color = enrichment)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    title = "Doublet Enrichment",
    x = "UMAP 1", y = "UMAP 2", color = "Enrichment"
  )

ggplot(doub_bmmc$originalDataUMAP, aes(x = X1, y = X2, color = score)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_color_viridis_c(option = "cividis") +
  theme_minimal() +
  labs(
    title = "Doublet Score (-log10 p)",
    x = "UMAP 1", y = "UMAP 2", color = "Score"
  )
```


## Creating ArchR project
```{r}
project_10XPBMC <- ArchRProject(
  ArrowFiles = ArrowFiles, 
  outputDirectory = "10XPBMC_Tutorial",
  copyArrows = FALSE
)
```

```{r}
project_10XPBMC
```
```{r}
project_10XPBMC@cellColData
```


```{r}
project_10XPBMC@cellColData

project_10XPBMC@peakSet
```


ArchR proj is small in size
```{r}
paste0("Memory Size = ", round(object.size(project_10XPBMC) / 10^6, 3), " MB")
```


```{r}
project_10XPBMC@cellColData
getCellColData(project_10XPBMC, select = "Sample")

getCellColData(ArchRProj = project_10XPBMC, select = "Sample")


getCellColData(project_10XPBMC, select = c("nFrags", "Sample"))

getCellColData(project_10XPBMC, select = c("log10(nFrags)", "nFrags - 1"))

df=getCellColData(project_10XPBMC, select = c("log10(nFrags)", "TSSEnrichment"))
```


```{r message=FALSE}
p <- ggPoint(
    x = df[,1], 
    y = df[,2], 
    colorDensity = TRUE,
    continuousSet = "sambaNight",
    xlabel = "Log10 Unique Fragments",
    ylabel = "TSS Enrichment",
    xlim = c(log10(500), quantile(df[,1], probs = 0.99)),
    ylim = c(0, quantile(df[,2], probs = 0.99))) + 
  geom_hline(yintercept = 4, lty = "dashed") + geom_vline(xintercept = 3, lty = "dashed")

p

#plotPDF(p, name = "TSS-vs-Frags.pdf", ArchRProj = project_10XPBMC, addDOC = FALSE)
```


```{r message=FALSE, warning=FALSE}
# ~ 1-2 mins
p_tss <- plotTSSEnrichment(ArchRProj = project_10XPBMC)
p_tss
```

## Filtering doublets
```{r}
project_10XPBMC <- filterDoublets(project_10XPBMC)

# Filtering 410 cells from ArchRProject!
#	scATAC_BMMC_R1 : 243 of 4932 (4.9%)
#	scATAC_CD34_BMMC_R1 : 107 of 3275 (3.3%)
#	scATAC_PBMC_R1 : 60 of 2453 (2.4%)
```

## Dimensionality Reduction and Clustering
```{r}
# ~ 2mins
project_10XPBMC <- addIterativeLSI(
    ArchRProj = project_10XPBMC,
    useMatrix = "TileMatrix", 
    name = "IterativeLSI")

project_10XPBMC@reducedDims$IterativeLSI$tileSize
```


```{r}
# ~ 10 secs
project_10XPBMC <- addClusters(
    input = project_10XPBMC,
    reducedDims = "IterativeLSI",
    method = "Seurat",
    name = "Clusters",
    resolution = 0.8, 
    seed = 112
)
```

```{r}
project_10XPBMC@cellColData
```


```{r}
table(project_10XPBMC$Clusters) # number of cells per cluster 
```


## Visualizing in a 2D UMAP Embedding
```{r}
project_10XPBMC <- addUMAP(
    ArchRProj = project_10XPBMC, 
    reducedDims = "IterativeLSI", 
    name = "UMAP", 
    nNeighbors = 30, 
    minDist = 0.5, 
    metric = "cosine",
    seed = 112
)

project_10XPBMC@embeddings$UMAP$df
```

```{r}
project_10XPBMC@embeddings$UMAP$df
```


```{r}
p1_umap <- plotEmbedding(ArchRProj = project_10XPBMC, 
                         colorBy = "cellColData", name = "Clusters", embedding = "UMAP")
p1_umap

p1_sample = plotEmbedding(ArchRProj = project_10XPBMC, 
                         colorBy = "cellColData", name = "Sample", embedding = "UMAP")
p1_sample

project_10XPBMC@cellColData

p1_nFrags = plotEmbedding(ArchRProj = project_10XPBMC, 
                         colorBy = "cellColData", name = "nFrags", embedding = "UMAP")
p1_nFrags
```

```{r}
plotPDF(p1_umap, p1_sample, p1_nFrags, name = "Plot-UMAP-Clusters.pdf", 
        ArchRProj = project_10XPBMC, addDOC = FALSE, width = 5, height = 5)
```

## You can also add t-SNE embeddings
```{r}
project_10XPBMC <- addTSNE(
    ArchRProj = project_10XPBMC, 
    reducedDims = "IterativeLSI", 
    name = "TSNE", 
    perplexity = 30,
    seed=112)

p1_tSNE <- plotEmbedding(ArchRProj = project_10XPBMC, 
                         colorBy = "cellColData", name = "Clusters", 
                         embedding = "TSNE")
p1_tSNE

plotPDF(p1_tSNE, name = "Plot-TSNE-Clusters.pdf", ArchRProj = project_10XPBMC, addDOC = FALSE, width = 5, height = 5)

p1_tSNE
```

```{r}
project_10XPBMC@embeddings$TSNE$df
```


## Assigning Clusters with Gene Scores
```{r}
project_10XPBMC <- addImputeWeights(project_10XPBMC)
```

```{r}
markerGenes  <- c(
    "CD34",  #Early Progenitor
    "GATA1", # Erythroid
    "PAX5", "MS4A1", "MME", #B-Cell Trajectory
    "CD14", "MPO", # Monocytes
    "CD3D", "CD8A"# T-Cells
  )

p_genes <- plotEmbedding(
    ArchRProj = project_10XPBMC, 
    colorBy = "GeneScoreMatrix", 
    name = markerGenes, 
    embedding = "UMAP",
    imputeWeights = getImputeWeights(project_10XPBMC)
)

class(p_genes)
p_genes$MS4A1
```

## Visualizing individual genes
```{r}
p_genes$MPO
```

## Visualizing all marker genes that were specified before
```{r}
#Rearrange for grid plotting
p_genes_all <- lapply(p_genes, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3),p_genes_all))
```

## Saving results as PDF
```{r}
plotPDF(plotList = p_genes, 
    name = "Plot-UMAP-Marker-Genes-W-Imputation.pdf", 
    ArchRProj = project_10XPBMC, 
    addDOC = FALSE, width = 5, height = 5)
```


## Visualizing Genome Browser Tracks
```{r}
p_tracks <- plotBrowserTrack(
    ArchRProj = project_10XPBMC, 
    groupBy = "Clusters", 
    geneSymbol = markerGenes, 
    upstream = 50000,
    downstream = 50000
)
```

```{r}
grid::grid.newpage()
grid::grid.draw(p_tracks$CD14)
```

```{r}
plotPDF(plotList = p_tracks, 
    name = "Plot-Tracks-Marker-Genes.pdf", 
    ArchRProj = project_10XPBMC, 
    addDOC = FALSE, width = 5, height = 5)
```


## Accessing interactive genome browser
```{r}
ArchRBrowser(ArchRProj = project_10XPBMC)
```


## Saving ArchR project for future use
```{r}
saveArchRProject(ArchRProj = project_10XPBMC, load = FALSE)
```

```{r}
sessionInfo()
```


# ------ THE END --------