---
title: "Template: Single-cell analysis using Seurat, part 2"
author: "Kevin Rue-Albrecht (modified by Lucy Garner)"
date: "2025-05-30"
date-format: long
format: html
---

```{r}
#| label: load-packages
#| include: false
library(Seurat)
library(tidyverse)
library(gprofiler2)
```

## Goal

Compare clustering results following standard "LogNormalize" normalisation and "SCTransform" normalisation.

## Exercise

### Import data

- Use `readRDS()` to import a preprocessed object from the file `seurat_after_qc.rds` provided.
- Name the object `seurat_after_qc`.

This object is the result of the Seurat workflow that we performed together yesterday.  
It is given to you again today to make sure we all start from the same point here.

```{r}

seurat_after_qc <- readRDS(file = "/project/shared/r/2_r_single_cell/2-seurat/seurat_after_qc.rds" )

```

- Display the Seurat object.

This is generally good practice, to keep a record of how your object looks at different stages of the analysis.

```{r}
seurat_after_qc
```

### Default assay

- Use `DefaultAssay()` to check the name of the active assay.

This is best practice before using Seurat functions that rely on the active assay.

```{r}
DefaultAssay(seurat_after_qc)
```

### SCTransform

- Use `SCTransform()` on the Seurat object to apply the SCTransform method.

**Hint:**

- Refer to the SCTransform vignette available at <https://satijalab.org/seurat/articles/sctransform_vignette.html>.  
- Consider confounding variables to specify in the argument `vars.to.regress = `.  
- Consider the number of variable features to specify in the argument `variable.features.n = `.

```{r}
seurat_after_qc <- SCTransform(seurat_after_qc, vars.to.regress = "percent_mt", variable.features.n = 2000)
```

**Note:** The `SCTransform()` function effectively replaces the workflow performed by the functions `NormalizeData()`, `ScaleData` and `FindVariableFeatures()`.

- Display the Seurat object again.

```{r}
seurat_after_qc
```

**What changed since you last displayed the object? Where are the normalised data stored?**

> Answer:  
>
> The number of features in the object has changed.
> It nearly doubled, which can be explained by the fact that there are now two assays in the object: 'RNA' and 'SCT'.
> 
> The active assay has changed from 'RNA' to 'SCT'.
> 
> 

**Note:** 

- The 'SCT' assay may have less features than the 'RNA' assay because `SCTransform()` calls `sctransform::vst()` with default argument `min_cells = 5`, which only retains features detected in at least 5 cells.

### Explore SCTransform results

- Combine `head()` and `[[]]` to display the cell metadata stored in the object.

This is good practice to keep track of information stored in the object.

```{r}
head(seurat_after_qc[[]])
```

**Can you see any new column(s) in the metadata?**

> Answer:  
> 
> 

### Visualise SCTransform results

- Use `VlnPlot()` to compare the total counts in the 'RNA' and 'SCT' assays.

**Hint:**

- Use the argument `same.y.lims = ` to force all y-axes to use the same range of values, to make comparisons easier.

```{r}
VlnPlot(object = seurat_after_qc, features = c('nCount_RNA', 'nCount_SCT'), same.y.lims = TRUE )

# SCT counts values meaningless, just creates by SCT analysis but shows that the values are on a more similar scale than the RNA counts

# plot_SCT <- VlnPlot(object = seurat_after_qc, features = 'nCount_RNA', assay = "SCT", layer = "counts" )
# 
# cowplot::plot_grid(plot_RNA, plot_SCT)
```

- Do the same thing to compare the number of features detected in the two assays.

```{r}
VlnPlot(object = seurat_after_qc, features = c('nFeature_RNA', 'nFeature_SCT'), same.y.lims = TRUE )
```

## Bonus

### Visualise a feature in the SCT assay

- For the feature of your choice (e.g., "LYZ"), using the "SCT" assay, compare the distribution of:
  - raw count data in the "counts" layer
  - log-normalised counts in the "data" layer
  - scaled log-normalised counts in the "scaled" layer

**Hint:**

- For each plot, pass the output of `FetchData()` to `ggplot()`.
- Store each plot as a separate object.
- Use `cowplot::plot_grid()` to combine all the plots into a single image.

```{r}
#| fig.height = 20
DefaultAssay(seurat_after_qc) <- "SCT"

ggplot_lyz_raw <- ggplot(FetchData(seurat_after_qc, vars = "LYZ", layer = "counts"), aes(LYZ)) +
  geom_histogram(fill = "grey", color = "black", bins = 100) +
  coord_cartesian(ylim = c(0, 500)) + 
  ggtitle("counts") +
  cowplot::theme_cowplot()

ggplot_lyz_normalised <- ggplot(FetchData(seurat_after_qc, vars = "LYZ", layer = "data"), aes(LYZ)) +
  geom_histogram(fill = "grey", color = "black", bins = 100) +
  coord_cartesian(ylim = c(0, 500)) +
  ggtitle("data") +
  cowplot::theme_cowplot()

ggplot_lyz_scaled <- ggplot(FetchData(seurat_after_qc, vars = "LYZ", layer = "scale.data"), aes(LYZ)) +
  geom_histogram(fill = "grey", color = "black", bins = 100) +
  coord_cartesian(ylim = c(0, 500)) +
  ggtitle("scaled") +
  cowplot::theme_cowplot()

cowplot::plot_grid(
  ggplot_lyz_raw,
  ggplot_lyz_normalised,
  ggplot_lyz_scaled, ncol = 1)
```

## Exercise

### Variable features

- Combine `VariableFeatures()` and `head()` to display the first 10 variable genes in the "SCT" assay.

**Hint:**

- For `VariableFeatures()`, use the `assay = ` argument to specify the assay to use.
- For `head()`, use the `n = ` argument to specify how many genes to display.

```{r}
head(VariableFeatures(object = seurat_after_qc, assay = "SCT"), n = 10)
```

- Repeat the same task for the "RNA" assay.

```{r}
head(VariableFeatures(object = seurat_after_qc, assay = "RNA"), n = 10)
```

- Combine the code above with `length()` and `intersect()` to display the number of genes that overlap the two subsets of variable genes.

```{r}
length(intersect(
  head(VariableFeatures(seurat_after_qc, assay = "SCT"), 10),
  head(VariableFeatures(seurat_after_qc, assay = "RNA"), 10)
))
```

**How many genes overlap? What is your takeaway message from it?**

> Answer:  
> 
> 

### Dimensionality reduction on SCT transformed data

- First, use `DefaultAssay()` to display the active assay (again!).

It is best practice to check that the active assay is correct before key steps when working with Seurat.  
Even better, it doesn't hurt to set the active assay to the desired one explicitly, to avoid any doubt.  

```{r}
DefaultAssay(seurat_after_qc)
```

- Use `RunPCA()` to run a principal component analysis on the Seurat object.

**Hint:**

- Set the argument `reduction.name = ` to something other than "pca", to avoid overwriting the existing PCA results based on the RNA assay. For instance, you could use "sct.pca".
  
```{r}
seurat_after_qc <- RunPCA(object = seurat_after_qc, assay = "SCT", reduction.name = "sct.pca" )
```

- Use `Reductions()` to display the name of the dimensionality reduction results present in the Seurat object at this point.

It is good practice to keep track of information stored in the object throughout the workflow.

```{r}
Reductions(seurat_after_qc)
```

### PCA elbow plot

- Use `ElbowPlot()` to produce a plot that shows the amount of variance explained by the top principal components.

**Hint:**

- Use the argument `ndims = ` to control the number of principal components displayed (up to the number of principal components computed earlier).
- Use the `reduction = ` argument to specify the dimensionality reduction result to use.
  
```{r}
ElbowPlot(object = seurat_after_qc, ndims = 50, reduction = "sct.pca")
```

**How many principal components would you use for downstream analyses?**

> Answer:  
> 20
> 

### UMAP

- Use `RunUMAP()` on the Seurat object, using the first 20 principal components computed earlier.

**Hint:**

- Use the argument `reduction = ` to specify the dimensionality reduction result to use.
- Set the argument `dims = ` to specify the principal components to use.
- Set the argument `reduction.name = ` to something other than "umap", to avoid overwriting the existing UMAP results based on the RNA assay. For instance, you could use "sct.umap".

```{r}
seurat_after_qc <- RunUMAP(object = seurat_after_qc, reduction = "sct.pca", dims = 1:20, reduction.name = "sct.umap")
```

- Use `DimPlot()` (twice) to display the two UMAP plots obtained using the RNA and SCT assays, respectively.

**Hint:**

- Use the `DimPlot()` function to produce each plot.
- Save each plot as a separate ggplot object.
- Use the `+` operator form the patchwork package or the function `cowplot::plot_grid()` to combine the ggplot objects into a single image.

```{r, fig.width = 10}
RNA_dimplot <- DimPlot(object = seurat_after_qc, reduction = "umap")

SCT_dimplot <- DimPlot(object = seurat_after_qc, reduction = "sct.umap")

cowplot::plot_grid(RNA_dimplot + SCT_dimplot )

```

### Clustering

- Use `FindNeighbors()` to compute the graph of nearest neighbours.

**Hint:**

- Set the argument `dims = ` to specify the principal components to use.
- Use the argument `reduction = ` to specify the dimensionality reduction result to use.
  
```{r}
seurat_after_qc <- FindNeighbors(object = seurat_after_qc, dims = 1:20, reduction = "sct.pca")
```

- Use `FindClusters()` to compute a cluster label for each cell.

**Hint:**

- Set the argument `resolution = ` to the value `0.5`. There is no right or wrong answer; we just want to have the same results here.

```{r}
seurat_after_qc <- FindClusters(object = seurat_after_qc, resolution = 0.5)
```

**Note:**

- If unspecified, the argument `graph.name = ` will concatenate the name of the active assay with '_snn' to determine the graph to use; [source code](https://github.com/satijalab/seurat/blob/c54e57d3423b3f711ccd463e14965cc8de86c31b/R/clustering.R#L433).
- Higher resolution, more clusters.

- For the record, use `Graphs()` to display the name of the graphs present in the Seurat object.

```{r}
Graphs(seurat_after_qc)
```

## Bonus

### Compare cluster labels using SCT and RNA assays

- Combine `colnames()` and `[[]]` to display the name of the columns of cell metadata present in the Seurat object.

```{r}
colnames(seurat_after_qc[[]])

```

- Use `table()` to compare the number of cells assigned to each cluster using the "SCT" and "RNA" assays, respectively.

**Hint:**

- Given two vectors of cluster labels, `table()` will display a table counting the number of cells assigned to each combination of cluster labels.
- Use the names of cell metadata columns displayed above to identify the two columns that contain the two sets of cluster labels.
- Name the arguments given to `table()`, for instance "RNA" and "SCT", to make the output easier to read and interpret.

```{r}
table(
  RNA_clusterLables = seurat_after_qc$RNA_snn_res.0.5, 
  SCT_clusterLabels = seurat_after_qc$SCT_snn_res.0.5 )
```

## Exercise

### Visualise cluster labels on the UMAP layout

- Use `DimPlot()` (twice) to colour cells by their cluster label on a UMAP layout, for results based on the 'RNA' and 'SCT' assay, respectively.

**Hint:**

- Use `DimPlot()` twice to produce the two separate plots for "RNA" and "SCT", respectively.
- Specifically, show the "RNA" clusters on the "RNA" UMAP layout, and the "SCT" clusters on the "SCT" UMAP layout.
- Use the argument `reduction = ` to specify the dimensionality reduction result to use in each plot.
- Use the argument `group.by = ` to specify the metadata column to colour each plot by.
- Optionally, use the argument `label = ` to add a label for each cluster in each plot.
- Optionally, use `ggtitle()` to add a title to each plot (to identify them more easily).

```{r}
RNA_clusterPlot <- DimPlot(object = seurat_after_qc, reduction = "umap", group.by = "RNA_snn_res.0.5", label = TRUE) + ggtitle("RNA_clusterPlot")

SCT_clusterPlot <- DimPlot(object = seurat_after_qc, reduction = "sct.umap", group.by = "SCT_snn_res.0.5", label = TRUE) + ggtitle("SCT_clusterPlot")

cowplot::plot_grid(RNA_clusterPlot +  SCT_clusterPlot)
```

**Based on this plot alone, do you have any preference for either the standard LogNormalize method or the SCTransform method?**

> Answer:  
> 
> The two methods largely agree with each other, suggesting that either one is acceptable.
> 
> The SCTransform method produces a couple of additional small clusters.
> This could be an advantage (higher sensitivity) or an issue (e.g., over-clustering).
> Only additional downstream analyses can answer that.
> 
> Visually, clusters edges seem cleaner in the UMAP produced by the SCT workflow.
> 
>

### Visualise gene expression on the UMAP layout

- Use `FeaturePlot()` to colour cells by the expression level of common cell type markers for PBMC for the UMAP layout based on the "SCT" assay.

**Hint:**

- Use the argument `features = ` to specify the names of the genes to colour cells by.
- Use the argument `reduction = ` to specify the name of the dimensionality reduction result to use in each plot.

```
- CD14 positive monocytes : LYZ, CD14
- CD16 monocytes          : FCGR3A, MS4A7
- CD4 positive T cells    : CD4, IL76
- CD8 positive T cells    : CD8A, CD3D
- NK cells                : GNLY, GZMB,NKG7
- B cells                 : MS4A1 , CD79A
- DC                      : CST3, FCER1A
- Platelets               : PPBP
```

```{r, fig.width = 10, fig.height = 8}
FeaturePlot(object = seurat_after_qc, features = c("LYZ", "CD14", "FCGR3A", "MS4A7", "CD4", "CD3D", "GNLY", "GZMB", "NKG7", "MS4A1", "CD79A", "CST3", "FCER1A", "PPBP"), reduction = "sct.umap" )
```

- Repeat the same plot for the UMAP layout based on the "RNA" assay.

```{r, fig.width = 10, fig.height = 8}
FeaturePlot(object = seurat_after_qc, features = c("LYZ", "CD14", "FCGR3A", "MS4A7", "CD4", "CD3D", "GNLY", "GZMB", "NKG7", "MS4A1", "CD79A", "CST3", "FCER1A", "PPBP"), reduction = "umap" )
```

**Do you think SCT improves the appearance of these particular plots?**

> Answer:  
> > Visually, the two layouts are fairly comparable.
> 

## Bonus

### Cluster markers (SCT)

- Use `FindAllMarkers()` to compute markers for the clusters obtained using the SCT assay.

**Hint:**

- Before using `FindAllMarkers()`, use `Idents()` to set the cell identities that should be used by `FindAllMarkers()`. `FindAllMarkers()` itself does not have an argument to specify cell identities.
- Use the argument `assay = ` to specify the assay to use ("RNA").
- Optionally, use the argument `only.pos = ` to request only positive markers (higher expression in the target cluster).
- Use the argument `min.pct = ` to request markers detected in at least 25% of cells in the target cluster.
- Set the argument `logfc.threshold = ` to 0.25, requesting markers showing a log fold-change greater than that value.

```{r}
Idents(seurat_after_qc) <- 'SCT_snn_res.0.5'

seurat_markers <- FindAllMarkers(object = seurat_after_qc, assay = "RNA", only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

```

### Over-representation analysis

- Use `gprofiler2::gost()` to perform pathway over-representation analysis (ORA) on cluster markers (i.e, genes differentially expressed between clusters) and visualise the results.

**Hint:**

- Prepare a named list where names are cluster labels and values are vectors of marker gene symbols for each cluster.
- First, combine `dplyr::group_by()`, `dplyr::filter()`, and `dplyr::select()` to produce a table of filtered markers.
- Specifically, filter marker genes for an adjusted p-value lower than 0.05 and an average log2 fold-change greater than `log2(1.5)` (i.e, a fold-change greater than 1.5).
- Then, use `split()` to make a list from a vector of cluster labels and gene names.
- Use `DefaultAssay()` to set the assay to "SCT", to avoid any doubt.
- Use `rownames()` to extract the gene symbols in the active assay of the Seurat object. Save those in an object that represents the background set of genes for the ORA analysis.
- For `gprofiler2::gost()`, use the argument `query = ` to provide the named list of markers, and the argument `custom_bg = ` to provide the background set of genes.

```{r}
seurat_markers_byCluster <- seurat_markers %>% 
  dplyr::group_by(cluster) %>%  
  dplyr::filter(p_val_adj < 0.05 & avg_log2FC > log2(1.5)) %>% 
  dplyr::select(cluster, gene)

marker_list <- split(x = seurat_markers_byCluster$gene,
  f = seurat_markers_byCluster$cluster)

DefaultAssay(seurat_after_qc) <- "SCT"

markers_background <- rownames(seurat_after_qc)
gost_sct <- gprofiler2::gost(query = marker_list, custom_bg = markers_background)
  
```

- Combine `head()` and `subset()` to display the first few biological pathways enriched in the markers of cluster "0".

```{r}
head(subset(gost_sct$result, query == "0" & source == "GO:BP"))
```

```{r}
sessionInfo()
```
